---
title: "sPLSDA batch correction with mice imputation"
output: html_document
date: "2024-12-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r}
rm(list = ls())
```

```{r}
output_dir = here("Samples538_D_ML/D3_Test_ext_data_BC/sPLSDA_Datasets")
```

# libraries General

```{r}
library(here)
library(tidyverse)
library(phyloseq)
library(cowplot)
library(ggplotify)
```

```{r}
# library(PCAtools) # for the command PCAtools::pca
library(mixOmics)
library(phyloseq)
library(microbiome)
library(PLSDAbatch)
library(vegan) #varpart

```

### Load Sample538 Maaslin 16S
```{r}
Samples538_Maaslin2_clr <- 
  read.table(here("Samples538_A_Geno_Microbiome/A2_Maaslin/Maaslin2_16S_Species_output/features",
                  "filtered_data_norm.tsv"),
                      header = T)
Samples538_Maaslin2_clr

Samples538_Maaslin2_clr = Samples538_Maaslin2_clr %>%
  rename(IID = feature) 
Samples538_Maaslin2_clr
```


## Load Sample538 metadata file
```{r}
Samples538_covariates <- 
  read.table(here("Samples538_A_Geno_Microbiome/A1_Plink/A1_00_Plink_files",
                           "Samples538_Combined_covariates_PCA.txt"),
                      header = T)
Samples538_covariates %>% head()

Samples538_covariates = Samples538_covariates %>% 
  dplyr::select(IID, ECC_status) %>% 
  rename(IID = IID) %>% 
  mutate(Study_name = "Khan_2024")
Samples538_covariates
```

## combine Sample538 Maaslin 16S and metadata
```{r}
Samples538_meta_Maaslin2_clr <- Samples538_covariates %>% 
  left_join(Samples538_Maaslin2_clr, by = "IID") %>% 
  # column_to_rownames(var = "IID") %>% 
  # factor the column ECC_status
  mutate(ECC_status = as.factor(ECC_status))
Samples538_meta_Maaslin2_clr
```


## Load external dataset
```{r}
# Define dataset names
dataset_names <- c("Agnello_2017", "Gomez_2017", "Kalpana_2020", "Teng_2015")  # External datasets

# Set the folder for the external datasets
dataset_folder <- here("Samples538_D_ML/D2_ML_ext_data/Datasets")  # Folder where the external dataset is stored

# Initialize empty lists to store the imported datasets
imported_datasets <- list()


# Loop through the dataset names
for (dataset_name in dataset_names) {
  
  ## Step 1: Import the primary dataset (.rds)
  dataset_path <- here(dataset_folder, paste0(dataset_name, "_MaAsLin2_CLR_ECCstatus.rds"))  
  dataset <- readRDS(dataset_path)
  
  # Add a new column with the study name
  dataset$Study_name <- dataset_name
  
  # Append the dataset to the list
  imported_datasets[[dataset_name]] <- dataset
  
}

# Combine all primary datasets
combined_ext_dataset_meta <- do.call(bind_rows, imported_datasets) %>% 
  relocate(Study_name, .after = ECC_status) %>%  # Move 'Study_name'
  rename(IID = sampleid) %>%  # Rename 'feature' to 'IID'
  identity()  # Ensure the dataset is a tibble
combined_ext_dataset_meta 
```


## imputation either mice pmm or 0

```{r, echo=TRUE, include=FALSE}
Pooled_Pl_Species_metadata <- bind_rows(Samples538_meta_Maaslin2_clr, combined_ext_dataset_meta) 

Pooled_Pl_Species_imp <- Pooled_Pl_Species_metadata %>%
  column_to_rownames(var = "IID") %>%  # Set the row names to the 'IID' column
  dplyr::select(-ECC_status, -Study_name) %>%  # Remove the 'Study_name' column
  as.data.frame() %>%                     # Ensure compatibility with mice
  # do either this
  mutate(across(everything(), ~ replace_na(.x, 0))) %>% # Replace NAs with 0
  # or this
  # mice::mice(method = "rf", m = 1, printFlag = TRUE) %>% mice::complete() %>%  # Apply PMM imputation
  rownames_to_column(var = "IID") %>%     # Re-add the row names as a column
  as_tibble()                             # Convert back to tibble for dplyr consistency


# Check the result
Pooled_Pl_Species_imp
# takes about 3 hours

# combine imputed data with metadata
Pooled_Pl_Species_imp_meta <- Pooled_Pl_Species_metadata[,c("IID", "ECC_status","Study_name")] %>% 
  inner_join(Pooled_Pl_Species_imp, by = "IID") %>% 
  identity()  # Ensure the dataset is a tibble
Pooled_Pl_Species_imp_meta %>% print()


# Save the object as .rds
saveRDS(Pooled_Pl_Species_imp_meta, file = here(output_dir,
                                "D3_01_Pooled_Pl_Species_imp_meta.rds"
                                )
        )
```

#Load

```{r}
# Load the object from .rds and assign it to a variable
Pooled_Pl_Species_imp_meta <- readRDS(file = here(output_dir,
                                "D3_01_Pooled_Pl_Species_imp_meta.rds")
)

# Verify the object
Pooled_Pl_Species_imp_meta
```

## Get the data for sPLSDA
```{r}
Pooled_Disease_status = as.factor(Pooled_Pl_Species_imp_meta$ECC_status)
Pooled_Disease_status
Pooled_Study_name = as.factor(Pooled_Pl_Species_imp_meta$Study_name)
Pooled_Study_name

Pooled_Pl_Species_RA_clr = Pooled_Pl_Species_imp_meta %>% 
  column_to_rownames(var = "IID") %>%
  dplyr::select( -ECC_status, -Study_name, )
Pooled_Pl_Species_RA_clr
```


```{r}
# # Check for zero-variance columns
# zero_variance_cols <- apply(Pooled_Pl_Species_RA_clr, 2, var) == 0
# 
# # Remove zero-variance columns
# Pooled_Pl_Species_RA_clr_filtZV <- Pooled_Pl_Species_RA_clr[, !zero_variance_cols]
```

#  pca option

```{r}
Pooled_Pl_Species_RA_OTU_mixOmics_pca_clr <- mixOmics::pca(Pooled_Pl_Species_RA_clr, scale = T,
                                                            center = T,
                                                            ncomp = 3,
                                                            )
```

```{r}
plot_before_splsda = 
mixOmics::plotIndiv(Pooled_Pl_Species_RA_OTU_mixOmics_pca_clr,
                    group=Pooled_Pl_Species_imp_meta$Study_name, 
                    ind.names = F,
                    legend=T,
                    point.lwd = 0.7,
                    pch = as.numeric(factor(Pooled_Pl_Species_imp_meta$ECC_status))+14,
                    pch.levels=(Pooled_Pl_Species_imp_meta$ECC_status), ellipse = TRUE, 
                    title="Before batch correction",legend.title = "Study_name", 
                    legend.title.pch = "ECC_status",
                    size.legend = rel(2.4),
                    size.legend.title = rel(1.0),
                    size.title = rel(1.2)
                    )
```
```{r}
plot_before_splsda_gg = plot_before_splsda$graph +theme(legend.position="none")
```


# check if the row names in metadata and OTU files are the same

```{r}
identical(row.names(Pooled_Pl_Species_imp_meta), row.names(Pooled_Pl_Species_RA_clr))
```

<!-- # Set batches and disease_status -->

<!-- ```{r} -->
<!-- Pooled_Disease_status = as.factor(Pooled_Pl_Species_imp_meta$Disease_status) -->
<!-- Pooled_Disease_status -->
<!-- Pooled_Study_name = as.factor(Pooled_Pl_Species_imp_meta$Study_name) -->
<!-- Pooled_Study_name -->
<!-- ``` -->


# The optimal number of components

```{r}
Pooled_Pl_Species_RA_clr_Disease_status.tune <- mixOmics::plsda(X = Pooled_Pl_Species_RA_clr, Y = Pooled_Disease_status, ncomp = 5)
Pooled_Pl_Species_RA_clr_Disease_status.tune$prop_expl_var #1
```

```{r}
Pooled_Pl_Species_RA_clr_Study_name.tune <- PLSDA_batch(X = Pooled_Pl_Species_RA_clr, 
                             Y.trt = Pooled_Disease_status, Y.bat = Pooled_Study_name,
                             ncomp.trt = 1, ncomp.bat = 10)
Pooled_Pl_Species_RA_clr_Study_name.tune$explained_variance.bat  #3 # the sum to top 2 Y values give a variance of 1 #see the next code 
```

```{r}
sum(Pooled_Pl_Species_RA_clr_Study_name.tune$explained_variance.bat$Y[1:4])
```

# To correct the Pooled_Pl_Species_RA_clr\_ data with PLSDA-batch

```{r}
Pooled_Pl_Species_RA_clr_plsda_batch.correct <- PLSDA_batch(X = Pooled_Pl_Species_RA_clr, 
                                      Y.trt = Pooled_Disease_status, Y.bat = Pooled_Study_name, 
                                      ncomp.trt = 1, ncomp.bat = 4)
Pooled_Pl_Species_RA_clr_plsda_batch <- Pooled_Pl_Species_RA_clr_plsda_batch.correct$X.nobatch
Pooled_Pl_Species_RA_clr_plsda_batch %>% as.data.frame()
```

```{r}
```

# sPLSDA-batch The optimal number of variables to select per component

```{r}
set.seed(789)
Pooled_Pl_Species_RA_clr_test.keepX = c(seq(1, 10, 1), seq(20, ncol(Pooled_Pl_Species_RA_clr), 10)) #number of OTUs
Pooled_Pl_Species_RA_clr_test.keepX
Pooled_Pl_Species_RA_clr_Disease_status.tune.v <- tune.splsda(X = Pooled_Pl_Species_RA_clr, Y = Pooled_Disease_status, 
                             ncomp = 1, test.keepX = Pooled_Pl_Species_RA_clr_test.keepX, 
                             validation = 'Mfold', folds = 4, nrepeat = 50)
Pooled_Pl_Species_splsda_choice_keepX = Pooled_Pl_Species_RA_clr_Disease_status.tune.v$choice.keepX 
Pooled_Pl_Species_splsda_choice_keepX
```

```{r}
Pooled_Pl_Species_RA_clr_splsda.correct <- PLSDA_batch(X = Pooled_Pl_Species_RA_clr, 
                                       Y.trt = Pooled_Disease_status, Y.bat = Pooled_Study_name, 
                                       ncomp.trt = 1, keepX.trt = Pooled_Pl_Species_splsda_choice_keepX, 
                                       ncomp.bat = 4)
Pooled_Pl_Species_RA_clr_splsda <- Pooled_Pl_Species_RA_clr_splsda.correct$X.nobatch
Pooled_Pl_Species_RA_clr_splsda %>%  as.data.frame()
```
# Save the sPLSDA batch corrected table
```{r}
# save(Pooled_Pl_Species_RA_clr_splsda, file = paste0(out_dir, "/Pooled_Pl_Species_RA_clr_splsda.RData"))
```


#  pca option

```{r}

Pooled_Pl_Species_RA_clr_splsda_pca <- mixOmics::pca(Pooled_Pl_Species_RA_clr_splsda, scale = T,
                                                            center = T,
                                                            ncomp = 3,
                                                            )
Pooled_Pl_Species_RA_clr_splsda_pca
```

```{r}
plot_after_splsda = 
mixOmics::plotIndiv(Pooled_Pl_Species_RA_clr_splsda_pca,
                    group=Pooled_Pl_Species_imp_meta$Study_name, 
                    ind.names = F,
                    legend=T,
                    point.lwd = 0.7,
                    pch = as.numeric(factor(Pooled_Pl_Species_imp_meta$ECC_status))+14,
                    pch.levels=(Pooled_Pl_Species_imp_meta$ECC_status), ellipse = TRUE, 
                    title="After batch correction",legend.title = "Study_name", 
                    legend.title.pch = "ECC_status",
                    size.legend = rel(2.4),
                    size.legend.title = rel(1.0),
                    size.title = rel(1.2)
                    )
```


```{r}


```


```{r}
plot_after_splsda_gg = plot_after_splsda$graph +theme(legend.position="none")
```

```{r}
legend = get_legend( plot_after_splsda$graph  + theme(legend.position="top"))

# add legend
p <- plot_grid(
  legend,
  plot_grid(plot_before_splsda_gg, plot_after_splsda_gg, ncol = 2),
  ncol = 1,
  rel_heights = c(0.15, 1) # Adjust these values to change the relative heights of the legend and plots
)
p
ggsave(here("Samples538_D_ML/D3_Test_ext_data_BC",
            "D3_01_before_after_splsda_gg.jpg"),
       dpi = 300,
       height = 5,
       width = 10)
```

```{r}
Pooled_Pl_Species_RA_clr_splsda_2 =
  Pooled_Pl_Species_RA_clr_splsda %>%
  as.data.frame() %>% 
  rownames_to_column("IID")
Pooled_Pl_Species_RA_clr_splsda_2
```
## Separate the individual tables for each external study
## merge the meta-data inclucing study name and disease status

```{r}
Pooled_Pl_Species_RA_clr_splsda_meta <- inner_join(
  Pooled_Pl_Species_imp_meta[, c("Study_name", "ECC_status", "IID")], # Ensure "IID" is included here for joining
  Pooled_Pl_Species_RA_clr_splsda_2,
  by = "IID" # Correct syntax for column matching
) %>%
  rename("sampleid" = "IID")

Pooled_Pl_Species_RA_clr_splsda_meta

```

```{r}

dataset_names <- c("Agnello_2017", "Gomez_2017", "Kalpana_2020", "Khan_2024", "Teng_2015")  # External datasets
# Iterate through each split
  for (dataset_name in dataset_names) {
    # Subset the data
    subset_data <- Pooled_Pl_Species_RA_clr_splsda_meta %>% 
      filter(Study_name == dataset_name) %>% 
      dplyr::select(-Study_name)

    # Define the file path
    file_path <- here(output_dir,
                                paste0(dataset_name, "_MaAsLin2_CLR_sPLSDA_ECCstatus.rds")
                      )
                           

    # Save the subset as an RDS file
    saveRDS(subset_data, here(output_dir,
                                paste0(dataset_name, "_MaAsLin2_CLR_sPLSDA_ECCstatus.rds"))
            )
    write.table(subset_data, file = here(output_dir,
                                paste0(dataset_name, "_MaAsLin2_CLR_sPLSDA_ECCstatus.txt")),
                sep = "\t", quote = F, row.names = F)
  }
```




























