---
title: "16S MRS comparison on ext CLR data"
output: html_document
date: "2024-01-02"
---

### MRS Score Calculation with Maaslin2 Coefficients and External Dataset Validation

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup
```{r}
library(dplyr)
library(here)
library(pROC)
library(tidyr)
library(ggplot2)
```

# Load Maaslin2 Coefficients and Filter for q-value < 0.01
```{r}
maaslin2_results <- readRDS(here("Samples538_A_Geno_Microbiome/A2_Maaslin",
                                 "A2_04_Species_coef_qval_16S_ITS.rds"))
maaslin2_results
# Filter significant species
significant_species <- maaslin2_results %>%
  filter(q_values < 0.01) %>%
  rename(Species = OTUs) %>%
  ## filter for non fungal species with F_s_ prefix
  filter(!grepl("F_s_", Species)) %>%
  dplyr::select(Species, Coefficients)

significant_species
```

1. A microbiome table


```{r}
# 16S data
Plaque538_16S_Species_CLR_OTU <- 
  read.table(here("Samples538_A_Geno_Microbiome/A2_Maaslin/Maaslin2_16S_Species_output/features",
                  "filtered_data_norm.tsv"),
                      header = T)
Plaque538_16S_Species_CLR_OTU

# ITS data
Plaque538_ITS_Species_CLR_OTU <- 
  read.table(here("Samples538_A_Geno_Microbiome/A2_Maaslin/Maaslin2_ITS_Species_output/features",
                  "filtered_data_norm.tsv"),
                      header = T)
## Combine the two tables
Plaque538_16S_ITS_Species_CLR_OTU = full_join(Plaque538_16S_Species_CLR_OTU,
                                              Plaque538_ITS_Species_CLR_OTU,
                                              by = "feature")
```


```{r}
Plaque538_16S_Species_CLR_OTU_2 = Plaque538_16S_Species_CLR_OTU %>%
  rename(IID = feature) %>% 
  replace(is.na(.), 0) %>%
  identity()
Plaque538_16S_Species_CLR_OTU_2

```

##covariate

## Load the metadata file
```{r}
Samples538_covariates <- 
  read.table(here("Samples538_A_Geno_Microbiome/A1_Plink/A1_00_Plink_files",
                           "Samples538_Combined_covariates_PCA.txt"),
                      header = T)
Samples538_covariates %>% head()

Samples538_covariates_2 = Samples538_covariates %>% 
  dplyr::select(IID, ECC_status) %>% #, Sex, Age, RuralUrban, SEFI_score, PC1, PC2, PC3, PC4, PC5) %>%
  mutate(across(c(ECC_status), as.factor)) %>% 
  # column_to_rownames(var = "IID") %>% 
  identity()

Samples538_covariates_2 

```


# combine all the datasets


```{r}
# Combine datasets by joining on "IID" and applying transformations in a streamlined manner
Samples538_covariates_vcf_clr_microbiome <- Samples538_covariates_2 %>%
  inner_join(Plaque538_16S_Species_CLR_OTU_2, by = "IID") %>%
  column_to_rownames(var = "IID") %>%
  # Remove rows with NA values
  # na.omit() %>%
  identity()

Samples538_covariates_vcf_clr_microbiome
```



# Prepare Train and Test Datasets
```{r}
# Load train dataset
train_data <- Samples538_covariates_vcf_clr_microbiome

# Ensure only significant species are selected
train_data_filtered <- train_data %>%
  dplyr::select(all_of(c("ECC_status", significant_species$Species)))

# Replace NA values with 0
train_data_filtered <- train_data_filtered %>% replace(is.na(.), 0)

# Split into train and test sets
set.seed(123)
split <- initial_split(train_data_filtered, prop = 0.8)
train_data <- training(split)
test_data <- testing(split)

train_data
test_data
```

# Calculate MRS Scores for Train and Test Datasets
```{r}
# Calculate MRS as the weighted sum of significant species abundances
mrs_train <- train_data %>%
  rowwise() %>%
  mutate(MRS_Score = sum(across(all_of(significant_species$Species)) * significant_species$Coefficient)) %>%
  ungroup()

mrs_test <- test_data %>%
  rowwise() %>%
  mutate(MRS_Score = sum(across(all_of(significant_species$Species)) * significant_species$Coefficient)) %>%
  ungroup()

mrs_train
mrs_test
```

# Evaluate Performance on Test Dataset
```{r}
# Determine the optimal threshold
roc_result_test <- roc(mrs_test$ECC_status, mrs_test$MRS_Score, levels = c("0", "1"), direction = "auto")
optimal_threshold <- coords(roc_result_test, "best", ret = "threshold", transpose = FALSE)

# Calculate AUROC
auc_test <- as.numeric(auc(roc_result_test))
auc_test

# Calculate AUPRC
auprc_test <- pr.curve(scores.class1 = mrs_test$MRS_Score[mrs_test$ECC_status == "0"], 
                       scores.class0 = mrs_test$MRS_Score[mrs_test$ECC_status == "1"], 
                       curve = FALSE)$auc.integral
auprc_test

# Add to results
test_results <- tibble(
  Dataset = "Khan_2024_test",
  AUROC = auc_test,
  AUPRC = auprc_test
)
```

# Apply MRS Equation on External Datasets and Calculate AUROC and AUPRC
```{r}
external_dataset_names <- c("Agnello_2017", "Gomez_2017", "Kalpana_2020", "Teng_2015")
external_results <- list()

for (dataset_name in external_dataset_names) {
  # Load external dataset
  dataset_folder <- here("Samples538_D_ML/D2_ML_ext_data/Datasets") 
  ext_data <- readRDS(file.path(dataset_folder,
                                paste0(dataset_name, "_MaAsLin2_CLR_ECCstatus.rds")))
  
  # Ensure columns match significant species
  ext_data <- ext_data %>%
    dplyr::select(any_of(c("ECC_status", significant_species$Species))) %>%
    replace(is.na(.), 0)
  
  # Calculate MRS for the external dataset
  ext_data <- ext_data %>%
    rowwise() %>%
    mutate(MRS_Score = sum(across(any_of(significant_species$Species)) * significant_species$Coefficient)) %>%
    ungroup()
  
  # Calculate AUROC for MRS scores
  roc_result <- roc(ext_data$ECC_status, ext_data$MRS_Score, levels = c("0", "1"), direction = "auto")
  auc_value <- as.numeric(auc(roc_result))  # Convert AUROC to numeric

  # Calculate AUPRC
  auprc_value <- pr.curve(scores.class1 = ext_data$MRS_Score[ext_data$ECC_status == "0"], 
                          scores.class0 = ext_data$MRS_Score[ext_data$ECC_status == "1"], 
                          curve = FALSE)$auc.integral

  external_results[[dataset_name]] <- tibble(Dataset = dataset_name, AUROC = auc_value, AUPRC = auprc_value)
}

# Combine results into a single dataframe
external_results_df <- bind_rows(external_results) %>%
  bind_rows(test_results)  # Add test dataset results

# Ensure "Khan_2024_test" is first
external_results_df <- external_results_df %>%
  mutate(Dataset = factor(Dataset, levels = c("Khan_2024_test", setdiff(Dataset, "Khan_2024_test"))))

# View results
external_results_df
```

# Plot AUROC and AUPRC Results for External Datasets
```{r}
external_results_long <- external_results_df %>%
  pivot_longer(cols = c(AUROC, AUPRC), names_to = "Metric", values_to = "Value")

ggplot(external_results_long, aes(x = Dataset, y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8, color = "black") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  labs(
    title = "MRS Performance on CLR values",
    x = "Dataset",
    y = "Model Pefromance",
    fill = "Metric"
  ) +
  scale_y_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) +
  theme_minimal(base_family = "Arial") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_text(face = "bold")
  )


# Save plot
ggsave(here("Samples538_D_ML/D2_ML_ext_data",
            "D3_03_MRS_Performance_CLR.png"),
       width = 5, height = 4)
```


