---
title: "C2_03_sig_variant_sig_OTU mediation on Samples538  with clr"
output: html_document
date: "2024-09-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# to run it in the background keep it commented out and run it in the terminal
```{r}
#  job::job({rmarkdown::render("/media/Second_stor/wasif/WK_Rprojects/RP_240910_Taste_Genes_Analysis_2/Samples538_C_Mediation/C2_MediationR/C2_02.Rmd")}, title = "Running Rmd C2_02 in background")
```


## R Markdown
```{r}
rm(list = ls())
```

```{r}
file_name_ext="C2_03_sig_variant_sig_OTU"
```

# Load libraries
```{r}
library(tidyverse)
library(here)
library(LDM)
library(ComplexHeatmap)
library(circlize)
library(grid)
```


## Need three tables
1. A microbiome table

16S microbiome
#### For microbiome data use Maaslin features for consistency in naming
```{r}
# Import the Maaslin features
Plaque538_16S_Species_CLR = readr::read_tsv(here("Samples538_A_Geno_Microbiome/A2_Maaslin/Maaslin2_16S_Species_output/features",
                                            "filtered_data_norm.tsv"))
Plaque538_16S_Species_CLR
```
ITS microbiome
```{r}
# Import the Maaslin features
Plaque538_ITS_Species_CLR = readr::read_tsv(here("Samples538_A_Geno_Microbiome/A2_Maaslin/Maaslin2_ITS_Species_output/features",
                                            "filtered_data_norm.tsv"))
Plaque538_ITS_Species_CLR

```


```{r}

Plaque538_16S_ITS_Species_CLR_OTU = full_join(Plaque538_16S_Species_CLR, Plaque538_ITS_Species_CLR, by = "feature") %>%
  rename(IID = feature)
Plaque538_16S_ITS_Species_CLR_OTU %>% head()

# Plaque538_16S_ITS_Species_CLR_OTU_2 = Plaque538_16S_ITS_Species_CLR_OTU %>%
#   # column_to_rownames(var = "IID") %>%
#   # dplyr::select(-FID) %>% 
#   identity()
# Plaque538_16S_ITS_Species_CLR_OTU_2 

Plaque538_16S_ITS_Species_CLR_OTU_names = Plaque538_16S_ITS_Species_CLR_OTU %>% 
  dplyr::select(-IID) %>%
  colnames()
Plaque538_16S_ITS_Species_CLR_OTU_names
```



## microbiome file with only Maaslin2 significant species

```{r}
Maaslin2_16S_ITS_features =
  readRDS(here("Samples538_A_Geno_Microbiome/A2_Maaslin",
               "A2_04_Species_coef_qval_16S_ITS.rds"))
Maaslin2_16S_ITS_features

# Maaslin2_16S_ITS_features_2 = Maaslin2_16S_ITS_features %>%
#   # column_to_rownames(var = "IID") %>%
#   dplyr::select(-FID) %>%
#   # as.matrix() %>%
#   # filter(IID %in% rownames(Samples538_covariates_vcf_raw)) %>%
#   column_to_rownames(var = "IID") %>%
#   identity()
# Maaslin2_16S_ITS_features_2 %>% head()

Maaslin2_16S_ITS_features_names = Maaslin2_16S_ITS_features %>%
  filter(q_values < 0.01) %>%
  pull(OTUs)
Maaslin2_16S_ITS_features_names

```




```{r}
# Plaque538_16S_ITS_Species_CLR_OTU_sig = Plaque538_16S_ITS_Species_CLR_OTU_2 %>%
#   # dplyr::select(one_of(Maaslin2_16S_ITS_features_names)) %>%
#   identity()
# Plaque538_16S_ITS_Species_CLR_OTU_sig

# ## feature names
# Plaque538_16S_ITS_Species_CLR_OTU_sig_names = Plaque538_16S_ITS_Species_CLR_OTU_sig %>%
#   dplyr::select(-IID) %>%
#   names()
# Plaque538_16S_ITS_Species_CLR_OTU_sig_names

```




## load recoded vcf file to make a new vcf which as only the above variants
```{r}
Samples538_recoded_vcf <- 
  read.table(here("Samples538_A_Geno_Microbiome/A1_Plink/A1_00_Plink_files",
                  "Samples538_55gene_Plink.vcf.2.raw"),
                        header = T)
Samples538_recoded_vcf[, 1:10]  %>% head()


# Create a vector of column names
columns_to_remove <- c("FID", "PAT", "MAT", "SEX", "PHENOTYPE")


Samples538_recoded_vcf_2 = Samples538_recoded_vcf %>%
  column_to_rownames(var = "IID") %>%
  dplyr::select(-c(columns_to_remove)) %>%
  # replace missing values with 0
  mutate(across(everything(), ~if_else(is.na(.), 0, .)))  %>% 
  # to make it a dominant vcf table
  mutate(across(everything(), ~case_when(
  . == 2 ~ 1,  # Replace 2 with 1
  TRUE ~ .     # Keep all other values the same
  ))) %>%
  # change the variants to factor
  mutate(across(everything(), as.factor)) %>%
  ## IID column is required for merging
  # rownames_to_column(var = "IID") %>% 
  identity()

Samples538_recoded_vcf_2 %>% head()
```
## Load the variant file
```{r}
Samples538_Plink_logit_covASPB_w_anno <- 
  read.table(here("Samples538_A_Geno_Microbiome/A1_Plink/A1_02_Plink_dominant",
                  "A1_02_Samples538_55gene_logistic_dominant_cov_anno.txt"),
                                                          header = T)
Samples538_Plink_logit_covASPB_w_anno %>% head()

Samples538_Plink_logit_covASPB_w_anno_2 = Samples538_Plink_logit_covASPB_w_anno %>% 
  arrange(P_BH) %>%                 # Sort by P_BH in ascending order
  slice_head(prop = 0.05) %>%          # Select the top 1% of rows
  filter(P_BH < 0.05) %>%
  filter(A1 != '*') %>%
  # mutate(SNP_A1 = paste(SNP, A1, sep = "_")) %>% 
  identity()
Samples538_Plink_logit_covASPB_w_anno_2 

Samples538_Plink_logit_covASPB_w_anno_2_sig = Samples538_Plink_logit_covASPB_w_anno_2 %>%
  filter(P_BH < 0.05) %>%
  pull(SNP)
Samples538_Plink_logit_covASPB_w_anno_2_sig
```

```{r}
Samples538_recoded_vcf_sig = Samples538_recoded_vcf_2 %>%
  dplyr::select(Samples538_Plink_logit_covASPB_w_anno_2$SNP) %>% 
  rownames_to_column(var = "IID")
Samples538_recoded_vcf_sig

## variant names
Samples538_recoded_vcf_sig_names = Samples538_recoded_vcf_sig %>% 
  dplyr::select(-IID) %>%
  names()
Samples538_recoded_vcf_sig_names 
```



##covariate

## Load the metadata file
```{r}
Samples538_covariates <- 
  read.table(here("Samples538_A_Geno_Microbiome/A1_Plink/A1_00_Plink_files",
                           "Samples538_Combined_covariates_PCA.txt"),
                      header = T)
Samples538_covariates %>% head()


```


```{r}
Samples538_covariates %>% group_by(Sex) %>% summarise(n = n())
Samples538_covariates %>% group_by(ECC_status) %>% summarise(n = n())
```

Required by plink Binary ('0' = control, '1' = case)
Required by plink Sex code ('1' = male, '2' = female, '0' = unknown)
FID (Family ID) and IID (Individual ID) for it to correctly identify individuals
mbQTL can not take non numerical covariates

```{r}

covariates <- c("Sex", "Age", "RuralUrban", "SEFI_score", "PC1", "PC2", "PC3", "PC4", "PC5")

Samples538_covariates_2 = Samples538_covariates %>% 
  dplyr::select(IID, ECC_status, Sex, Age, RuralUrban, SEFI_score, PC1, PC2, PC3, PC4, PC5) %>%
  # mutate(FID = IID) %>%
  mutate(Sex = case_when( # AS Original in data 1, Female | 2, Male 
    Sex == 0 ~ "Female",
    Sex == 1 ~ "Male")) %>% 
  mutate(Sex = case_when( # AS 	required by plink Sex code ('1' = male, '2' = female, '0' = unknown)
    Sex == "Female" ~ 2,
    Sex == "Male" ~ 1,
    TRUE ~ 0 )) %>% 
  dplyr::select(IID, ECC_status, covariates) %>%
  mutate(across(c(ECC_status, Sex, RuralUrban), as.factor)) %>% 
  # column_to_rownames(var = "IID") %>% 
  identity()

Samples538_covariates_2 


# Samples538_covariates_2_t = Samples538_covariates_2 %>%
#   as.matrix() %>% 
#   t() %>% 
#   as.data.frame()
# Samples538_covariates_2_t 
```










# combine all the datasets


```{r}
# Combine datasets by joining on "IID" and applying transformations in a streamlined manner
Samples538_covariates_vcf_clr_microbiome <- Samples538_covariates_2 %>%
  inner_join(Samples538_recoded_vcf_sig, by = "IID") %>%
  inner_join(Plaque538_16S_ITS_Species_CLR_OTU, by = "IID") %>%
  # Remove rows with NA values
  na.omit() %>%
  identity()

Samples538_covariates_vcf_clr_microbiome
```




```{r}
library(mediation)
mediator_model <- glm(B_s_Streptococcus_mutans ~ chr15_40303364_PLCB2_rs2305645 +Age+Sex+PC1+PC2+PC3+PC4+PC5+RuralUrban+SEFI_score,
                     data =Samples538_covariates_vcf_clr_microbiome,
                     family = gaussian)

outcome_model <- glm(ECC_status ~ chr15_40303364_PLCB2_rs2305645 + B_s_Streptococcus_mutans +Age+Sex+PC1+PC2+PC3+PC4+PC5+RuralUrban+SEFI_score,
                    data = Samples538_covariates_vcf_clr_microbiome,
                    family = binomial(link = "logit"))

results <- mediate(mediator_model,
                   outcome_model,
                   covariates = c("Age", "Sex", "PC1", "PC2", "PC3", "PC4", "PC5", "RuralUrban", "SEFI_score"),
                   treat= 'chr15_40303364_PLCB2_rs2305645',
                   mediator= 'B_s_Streptococcus_mutans' ,
                   boot=TRUE,
                   sims=2)
summary(results, digits = 20) 
    # Extract ACME (average) estimate and p-value
    results$d.avg
    results$d.avg.p

    # Extract ADE (average) estimate and p-value
    results$z.avg
    results$z.avg.p

    # Extract Total Effect (TE) estimate and p-value
    results$tau.coef
    results$tau.p
```

```{r}
results$d.avg.p

```

```{r}

# Create an empty dataframe to store the results
results_df <- data.frame(
  variant = character(),
  OTU = character(),
  acme_estimate = numeric(),
  acme_p_value = numeric(),
  ade_estimate = numeric(),
  ade_p_value = numeric(),
  total_effect_estimate = numeric(),
  total_effect_p_value = numeric(),
  stringsAsFactors = FALSE
)

# Iterate over each combination of variant and OTU
for (OTU in Maaslin2_16S_ITS_features_names) {
  for (variant in Samples538_recoded_vcf_sig_names) {

    # Fit the mediator model
    mediator_model <- glm(as.formula(paste0(OTU, " ~ ", variant,
                                            " + Age + Sex + PC1 + PC2 + PC3 + PC4 + PC5 + RuralUrban + SEFI_score")),
                          data = Samples538_covariates_vcf_clr_microbiome,
                          family = gaussian)

    # Fit the outcome model
    outcome_model <- glm(as.formula(paste0("ECC_status ~ ", variant, " + ", OTU,
                                           " + Age + Sex + PC1 + PC2 + PC3 + PC4 + PC5 + RuralUrban + SEFI_score")),
                         data = Samples538_covariates_vcf_clr_microbiome,
                         family = binomial(link = "logit"))

    # Perform mediation analysis
    med_results <- mediate(mediator_model,
                           outcome_model,
                           treat = variant,
                           mediator = OTU,
                           sims = 1000)

    results_df <- results_df %>%
      add_row(
              variant = variant,
              OTU = OTU,
              acme_estimate = med_results$d.avg,
              acme_p_value = med_results$d.avg.p,
              ade_estimate = med_results$z.avg,
              ade_p_value = med_results$z.avg.p,
              total_effect_estimate = med_results$tau.coef,
              total_effect_p_value = med_results$tau.p
              )
  }
}

```

```{r}
# Display the results dataframe
print(results_df)
```

# Save the results to a file
```{r}

write.table(results_df,
          file = here("Samples538_C_Mediation/C2_MediationR",
                      paste0(file_name_ext, "_mediationR.csv")),
          row.names = FALSE, sep = "\t", quote = FALSE)

```


```{r}
rm(results_df)
```


# Load the results
```{r}
results_df <- read.table(here("Samples538_C_Mediation/C2_MediationR",
                      paste0(file_name_ext, "_mediationR.csv")),
                         header = TRUE, sep = "\t")
results_df
```



```{r}

results_df_2 = results_df %>% 
  # replace acme_p_value == 0 to 10^-16 accoring to the minimumn threshold
  mutate(adjusted_p_value = ifelse(acme_p_value == 0, 10^-16, acme_p_value)) %>%
  # Group by 'variant' and adjust p-values within each group
  group_by(variant) %>%
  mutate(adjusted_p_value = p.adjust(acme_p_value, method = "fdr")) %>%
  ungroup()  # Ungroup after the operation

# Display the updated results
print(results_df_2)
  
```
```{r}
results_df_3 = results_df_2 %>% 
  # Filter out rows with adjusted p-value > 0.05
  filter(adjusted_p_value <= 0.05) %>%
  # Arrange the rows by adjusted p-value
  arrange(adjusted_p_value)
results_df_3
```
```{r}
# Convert the adjusted p-values to -log10 scale for better visualization
results_df_3 <- results_df_3 %>%
  mutate(log_adjusted_p_value = -log10(adjusted_p_value)) %>% 
  complete(variant, OTU, fill = list(log_adjusted_p_value = 0))  # Add missing combinations

# Create the heatmap
heatmap_plot <- ggplot(results_df_3, aes(x = variant, y = OTU, fill = log_adjusted_p_value)) +
  geom_tile(color = "black") +
  scale_fill_gradientn(colors = c("white","yellow", "yellow2", "yellow3","yellow4", "pink","pink2", "pink3","pink3","red", "red2", "red3", "darkred"),
                       name = "-log10(adjusted p-value)") +
  labs(x = "Variant", y = "OTU",
       # title = "Heatmap of Adjusted p-values"
       ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank()
) 

# Display the heatmap
print(heatmap_plot)

ggsave(here("Samples538_C_Mediation/C2_MediationR",
         paste0(file_name_ext, "_Heatmap_Mediation.jpg")),
       plot = heatmap_plot,
       width = 6,
       height = 4)
```

