---
title: "Plink with Age,Sex, model"
output: html_document
date: "2024-08-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r}
rm(list = ls())

```

# Load libraries
```{r}
library(tidyverse)
library(here)
library(snpStats)
library(vcfR)
library(mbQTL)
library(RColorBrewer)
library(qqman)
```

# create plink files

```{bash}
source /home/wasif_pclab/anaconda3/etc/profile.d/conda.sh

conda activate Plink_env
new_Plink_dir=/media/Second_stor/wasif/WK_Rprojects/RP_240910_Taste_Genes_Analysis_2/Samples538_A_Geno_Microbiome/A3_AIC_BIC_comp/A3_Plink_files/
rm -rf  $new_Plink_dir
mkdir   $new_Plink_dir
cd      $new_Plink_dir

metadata_dir=/media/Second_stor/wasif/WK_Rprojects/RP_230718_Taste_Genes_Analysis_1/Samples538_Analysis/Metadata538_files/
inupt_dir=/media/Second_stor/wasif/WK_Rprojects/RP_230718_Taste_Genes_Analysis_1/Samples538_Analysis/Samples538_GATK/

plink \
--vcf ${inupt_dir}/Samples538_55genes_GenotypeGVCFs_SNPIDs_filtered_anno_IDcomplete5.vcf \
--double-id \
--make-bed \
--keep-allele-order \
--make-pheno ${metadata_dir}/Samples538_Combined_Phenotype.txt 1 \
--update-sex ${metadata_dir}/Samples538_Combined_Phenotype_with_sex_info.txt \
--recode \
--dummy-coding 'no-round' \
--covar ${metadata_dir}/Samples538_Combined_covariates.txt \
--out ./Samples538_55gene_Plink
```



# Plink recode vcf
```{bash}
source /home/wasif_pclab/anaconda3/etc/profile.d/conda.sh

conda activate Plink_env

cd       /media/Second_stor/wasif/WK_Rprojects/RP_240910_Taste_Genes_Analysis_2/Samples538_A_Geno_Microbiome/A3_AIC_BIC_comp/A3_Plink_files/

plink --file ./Samples538_55gene_Plink \
--recode A \
--out ./Samples538_55gene_Plink.vcf
```


###  Generate the PCA components:
```{bash}
source /home/wasif_pclab/anaconda3/etc/profile.d/conda.sh

conda activate Plink_env
cd /media/Second_stor/wasif/WK_Rprojects/RP_240910_Taste_Genes_Analysis_2/Samples538_A_Geno_Microbiome/A3_AIC_BIC_comp/A3_Plink_files/
plink \
--bfile ./Samples538_55gene_Plink \
--geno 0.01 \
--mind 0.5 \
--maf 0.01 \
--hwe 0.00001 \
--pca 5 \
--out ./Samples538_55gene_Plink_PCA

## PCs can not be calculated without mind option
```

```{r}
# # Load eigenvalues from the file
# eigenvalues <- read.table(here("Samples538_A_Geno_Microbiome/A3_AIC_BIC_comp/A3_Plink_files/",
#                                "Samples538_55gene_Plink_PCA.eigenval"),
#                           header = FALSE, as.is = TRUE)$V1
# 
# # Create a tibble for plotting
# eigen_df <- tibble(
#   Principal_Component = paste0("PC", 1:length(eigenvalues)),
#   Eigenvalue = eigenvalues
# )
# 
# # Create the scree plot
# # Create the scree plot with bars, points, lines, and custom y-axis limits
# ggplot(eigen_df, aes(x = Principal_Component, y = Eigenvalue)) +
#   geom_bar(stat = "identity", fill = "skyblue", alpha = 0.7) +  # Adding bars
#   geom_point(size = 4) +
#   geom_line(group = 1, color = "blue", linetype = "dashed") +
#   ylim(0, 100) +  # Setting y-axis limits from 0 to 100
#   theme_minimal() +
#   labs(title = "Scree Plot",
#        x = "Principal Component",
#        y = "Variance explained (%)") +
#   theme(text = element_text(size = 14))
```

Merge the PCA results with your phenotype file:
Import metadata file
```{r}
pca_data = here("Samples538_A_Geno_Microbiome/A3_AIC_BIC_comp/A3_Plink_files",
                 "Samples538_55gene_Plink_PCA.eigenvec") %>%
  read.table(header = FALSE, sep = "", comment.char = "#", as.is = TRUE) %>%
  as_tibble() %>% 
  rename(FID = V1, IID = V2, PC1 = V3, PC2 = V4, PC3 = V5, PC4 = V6, PC5 = V7)
pca_data %>% head()


Samples538_metadata <-here("Samples538_A_Geno_Microbiome/A3_AIC_BIC_comp/A3_Plink_files",
                           "Samples538_55gene_Plink.cov") %>% 
                                  read.table( header = T, as.is = TRUE)
Samples538_metadata %>% head()

## add PC1 and PC2 to metadata by IID and FID
Samples538_metadata <- left_join(Samples538_metadata,
                                 pca_data,
                                 by = c("FID", "IID"))
Samples538_metadata %>% head()

write.table(Samples538_metadata,
            here("Samples538_A_Geno_Microbiome/A3_AIC_BIC_comp/A3_Plink_files",
                 "Samples538_Combined_covariates_PCA.txt"),
            quote = F,
            sep = "\t",
            row.names = F)

```



# plink logistic association and chi square test
# Chi square test is required to get hte F_A and F_U values

```{bash}
source /home/wasif_pclab/anaconda3/etc/profile.d/conda.sh

conda activate Plink_env

cd /media/Second_stor/wasif/WK_Rprojects/RP_240910_Taste_Genes_Analysis_2/Samples538_A_Geno_Microbiome/A3_AIC_BIC_comp/A3_Plink_files/

metadata_dir=/media/Second_stor/wasif/WK_Rprojects/RP_230718_Taste_Genes_Analysis_1/Samples538_Analysis/Metadata538_files/

plink \
--bfile ./Samples538_55gene_Plink \
--geno 0.01 \
--maf 0.01 \
--hwe 0.00001 \
--assoc \
--ci 0.95 \
--make-pheno ${metadata_dir}/Samples538_Combined_Phenotype.txt 1 \
--out ./Samples538_55gene_Plink_geno.01_maf.01_hwe.00001_assoc


plink \
--bfile ./Samples538_55gene_Plink \
--geno 0.01 \
--maf 0.01 \
--hwe 0.00001 \
--logistic \
--ci 0.95 \
--covar ./Samples538_Combined_covariates_PCA.txt \
--covar-name Age,Sex  \
--make-pheno ${metadata_dir}/Samples538_Combined_Phenotype.txt 1 \
--out ./Samples538_55gene_Plink_geno.01_maf.01_hwe.00001_logistic
```

# add genes to the plink assoc file and 

```{r}

# Read VCF file
vcf <- read.vcfR(here("../RP_230718_Taste_Genes_Analysis_1/Samples538_Analysis/Samples538_GATK/",
                      "Samples538_55genes_GenotypeGVCFs_SNPIDs_filtered_anno_IDcomplete5.vcf"))


# Extract chromosome and position
chrom <- vcf@fix[,1]
pos <- vcf@fix[,2]

# Extract the INFO column
info <- vcf@fix[,8]

# Extract the ANN information from INFO by removing the 'ANN=' prefix
ann_info <- gsub("^ANN=", "", info)

# Each element of 'ann_info' can have multiple annotations (separated by comma),
# and each annotation has multiple sub-fields (separated by '|').

# Function to parse a single annotation
# Updated parse_ann function from the previous correction
parse_ann <- function(ann) {
  # Split the annotation into sub-fields
  sub_fields <- strsplit(ann, "\\|")[[1]]

  # Check if there are enough sub-fields to extract the desired information
  if (length(sub_fields) >= 11) {
    gene_name <- sub_fields[4]
    variant_type <- sub_fields[2]
    hgvs_p <- sub_fields[11] # Extract HGVS.p information
    return(c(gene_name, variant_type, hgvs_p))
  } else {
    return(c(NA, NA, NA))
  }
}

rm(vcf)
# Parse the ANN field for each variant to include HGVS.p information
genes_and_variant_types  <- t(sapply(ann_info, parse_ann)) # transpose to make each row contain the gene, variant type, and HGVS.p

# Modify the data frame creation to include HGVS.p information
df_chromPOS_to_gene_variant  <- data.frame(
  CHROM = chrom,
  POS = pos,
  Gene = genes_and_variant_types [,1],
  Variant_Type = genes_and_variant_types [,2],
  Variant_Effect = genes_and_variant_types [,3] # Include the HGVS.p information
)

df_chromPOS_to_gene_variant %>% head()

df_chromPOS_to_gene_variant =  df_chromPOS_to_gene_variant %>%
  # we do not need to print the effect of synonymous_variant as it is the same as the reference
  mutate(Variant_Effect = if_else(Variant_Type == "synonymous_variant", "", Variant_Effect)) %>%
  # replace those values as empty cells have problem in saving as word file
 mutate(Variant_Effect = if_else(Variant_Effect == "", "-", Variant_Effect)) %>% 
  mutate(CHROM = gsub("chr", "", CHROM))
df_chromPOS_to_gene_variant %>% head()
# genes_and_variant_types <- t(sapply(ann_info, parse_ann)) # transpose to make each gene-variant pair a row
#
# # Create a dataframe with the chromosome, position, gene name, and variant type
# df_chromPOS_to_gene_variant <- data.frame(CHROM = chrom, POS = pos, Gene = genes_and_variant_types[,1], VARIANT_TYPE = genes_and_variant_types[,2])
# df_chromPOS_to_gene_variant
```

## Assign gene name to assoc file
```{r}
logit_assoc_file = read.table(here("Samples538_A_Geno_Microbiome/A3_AIC_BIC_comp/A3_Plink_files",
                             "Samples538_55gene_Plink_geno.01_maf.01_hwe.00001_logistic.assoc.logistic"),
                        header = T)
logit_assoc_file = logit_assoc_file %>%
  mutate( BP = as.character(BP),
         CHR = as.character(CHR)) %>% 
  filter(TEST == "ADD") 
logit_assoc_file %>% head()
```

```{r}
chi_assoc_file = read.table(here("Samples538_A_Geno_Microbiome/A3_AIC_BIC_comp/A3_Plink_files",
                             "Samples538_55gene_Plink_geno.01_maf.01_hwe.00001_assoc.assoc"),
                        header = T)
chi_assoc_file = chi_assoc_file %>%
  mutate( BP = as.character(BP),
         CHR = as.character(CHR)) %>% 
  dplyr::select(CHR, SNP, BP, A1, A2, F_A, F_U) 
chi_assoc_file %>% head()
```


## To select only those genes which are in bed file
```{r}
Gene55_names = read.table(here("../RP_230718_Taste_Genes_Analysis_1/References_Seqs/bed_files",
                               "Gene55_names_only.txt"),
                        header = F)
Gene55_names
```


# merge assoc file with gene names and chi sqare results

```{r}
assoc_file_df_chromPOS_to_gene_variant =  inner_join(logit_assoc_file, df_chromPOS_to_gene_variant,
                                             by = c("CHR" = "CHROM", "BP" = "POS")) %>% 
  ## To add the information related to F_A and F_U
  inner_join(., chi_assoc_file, by = c("CHR" = "CHR",
                                    "BP" = "BP",
                                    "SNP" = "SNP",
                                    "A1" = "A1"
                                    ))
assoc_file_df_chromPOS_to_gene_variant %>% head()

assoc_file_df_chromPOS_to_gene_variant_adjP = assoc_file_df_chromPOS_to_gene_variant %>% 
  mutate(P_BH = p.adjust(P, method = "BH"),
         P_Bonferroni = p.adjust(P, method = "bonferroni")  ) %>% 
  arrange(P_Bonferroni, SNP) %>% 
  mutate(BP= as.numeric(BP)) %>% 
  # to filter only those genes which are in bed file
  filter(Gene %in% Gene55_names$V1)
assoc_file_df_chromPOS_to_gene_variant_adjP %>% head()

```



```{r}
assoc_file_df_chromPOS_to_gene_variant_adjP_3 = assoc_file_df_chromPOS_to_gene_variant_adjP %>% 
  # arrange(Gene) %>% 
  mutate(Location_GRCh38 = paste0("chr", CHR, ":", BP),
         Frq_ECC = F_A,
         Frq_CF = F_U) %>% 
  dplyr::select(Gene, SNP, Location_GRCh38, A1, A2,  STAT, P, OR,
                P_Bonferroni, Variant_Type,Variant_Effect,
                Frq_ECC, Frq_CF, SE) %>% 
  # extract SNP rsID
  separate(SNP, into = c(NA, NA, NA, "SNP_rsID"), sep = "_+", remove = FALSE)
assoc_file_df_chromPOS_to_gene_variant_adjP_3

write.table(assoc_file_df_chromPOS_to_gene_variant_adjP_3, 
            here("Samples538_A_Geno_Microbiome/A3_AIC_BIC_comp",
                 "A3_01_Samples538_Plink_logit_covASPB_w_anno.txt"),
            sep = "\t", quote = F, row.names = F)

```

<!-- ## Load the file -->
<!-- ```{r} -->
<!-- Samples538_Plink_logit_covASPB_w_anno = read.table(here("Samples538_A_Geno_Microbiome/A3_AIC_BIC_comp", -->
<!--                                                                "A3_01_Samples538_Plink_logit_covASPB_w_anno.txt"), -->
<!--                                                           header = T) -->
<!-- Samples538_Plink_logit_covASPB_w_anno %>% head() -->
<!-- ``` -->


<!-- # QQ plot -->
<!-- ```{r} -->
<!-- # Assuming your dataframe is named 'Samples538_Plink_logit_covASPB_w_anno' -->
<!-- # Extract the P-values -->
<!-- p_values <- Samples538_Plink_logit_covASPB_w_anno$P -->

<!-- # Create the Q-Q plot -->
<!-- qq(p_values, main = "Q-Q Plot of GWAS P-values") #%>%  -->
<!--   # ggplotify::as.ggplot() -->

<!-- ``` -->

<!-- ## ggplot  -->
<!-- ```{r} -->
<!-- ggplot_title = "Manhattan Plot with logistic regression" -->
<!-- ggplot_subtitle = "Plink geno.01_maf.01_hwe.00001 \nCovariates: Age Sex PC1-5 SEFI_score Urban_status" -->
<!-- filext = "A3_01_Samples538_55gene_" -->
<!-- ``` -->


<!-- ## QQ plot in ggplot for the p-values -->

<!-- ```{r} -->
<!-- # Ensure your dataframe has no NA values in the P column -->
<!-- Samples538_Plink_logit_covASPB_w_anno_qq <- Samples538_Plink_logit_covASPB_w_anno[!is.na(Samples538_Plink_logit_covASPB_w_anno$P), ] -->

<!-- # Calculate expected and observed quantiles -->
<!-- observed <- -log10(sort(Samples538_Plink_logit_covASPB_w_anno_qq$P)) -->
<!-- expected <- -log10(ppoints(length(Samples538_Plink_logit_covASPB_w_anno_qq$P))) -->

<!-- # Create a data frame for plotting -->
<!-- qq_data <- data.frame(observed = observed, expected = expected) -->

<!-- # Calculate the genomic inflation factor (lambda) -->
<!-- lambda <- median(qchisq(1 - Samples538_Plink_logit_covASPB_w_anno_qq$P, 1)) / qchisq(0.5, 1) -->

<!-- # Create the plot -->
<!-- ggplot(qq_data, aes(x = expected, y = observed)) + -->
<!--   geom_point(alpha = 0.6) + -->
<!--   geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") + -->
<!--   labs(x = expression("Expected -log10("*italic(P)*")"),  -->
<!--      y = expression("Observed -log10("*italic(P)*")"), -->
<!--        # title = "Q-Q Plot of GWAS P-values", -->
<!--        subtitle = paste("Genomic inflation factor (λ) =", round(lambda, 2)) -->
<!--        ) + -->
<!--   theme_bw() + -->
<!--   theme(plot.title = element_text(hjust = 0.5), -->
<!--         plot.subtitle = element_text(hjust = 0.5))  -->

<!-- ggsave (here("Samples538_A_Geno_Microbiome/A3_AIC_BIC_comp", -->
<!--               paste0(filext,"QQ_plot.png")), width = 6, height = 6, dpi = 300) -->
<!-- ``` -->


<!-- ## Use the data to make a volcano plot -->

<!-- ```{r} -->
<!-- # Assume assoc_file_df_chromPOS_to_gene_variant_adjP is your dataset -->
<!-- volcano_data = Samples538_Plink_logit_covASPB_w_anno %>% -->
<!--   mutate(log10p = -log10(P), logOR = log2(OR)) %>% -->
<!--   arrange(P) %>% -->
<!--   filter(!is.na(log10p)) %>% -->
<!--   mutate(Significance = ifelse(log10p >= 3, "Significant", "Not Significant")) -->

<!-- # Define shapes and variant order combined into one step -->
<!-- shape_mapping <-  c( -->
<!--   "conservative_inframe_deletion" = 13, -->
<!--   "synonymous_variant" = 17, -->
<!--   "missense_variant" = 18, -->
<!--   "intron_variant" = 16, -->
<!--   "frameshift_variant" = 11, -->
<!--   "downstream_gene_variant" = 9, -->
<!--   "Other" = 4) -->
<!-- # Dynamically generate variant order from shape_mapping -->
<!-- variant_order <- names(shape_mapping) -->

<!-- # Create a new column in volcano_data to classify unspecified variants as "Other" -->
<!-- volcano_data <- volcano_data %>% -->
<!--   mutate(Variant_Type_Labeled = -->
<!--            factor(ifelse(Variant_Type %in% names(shape_mapping),Variant_Type, "Other"),  -->
<!--                                        levels = variant_order)) -->


<!-- # Determine axis limits for symmetry, ensuring it includes your specified marks -->
<!-- max_abs_logOR <- max(c(abs(volcano_data$logOR), 4), na.rm = TRUE) -->



<!-- # Create the plot with improved label placement using ggrepel -->
<!-- volcano_plot <- ggplot(volcano_data, aes(x = logOR, y = log10p)) + -->
<!--   geom_point(aes(color = Significance, shape = Variant_Type_Labeled), alpha=0.8) + -->
<!--   scale_shape_manual(values = shape_mapping) + # Ensure "Other" is included -->

<!--   scale_color_manual(values = c("Not Significant" = "black", "Significant" = "red")) + -->
<!--   labs(title = ggplot_title, -->
<!--        subtitle = ggplot_subtitle, -->
<!--        caption = "-", -->
<!--        x = "Log Odds Ratio", y = expression("-log10"~italic(P)~"value")) + -->
<!--   theme_bw() + -->
<!--   geom_hline(yintercept = 3, linetype = "dashed", color = "black") + -->
<!--   geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") + -->
<!--   ggrepel::geom_label_repel(data = filter(volcano_data, Significance == "Significant"), aes(label = Gene), -->
<!--                   box.padding = 0.1, point.padding = 0.1, -->
<!--                   nudge_x = 0, nudge_y = 0, -->
<!--                   segment.color = 'black',  # Ensure segment color is visible -->
<!--                   arrow = NULL, -->
<!--                   min.segment.length = 0) + # Adding arrowheads to lines -->
<!--   scale_x_continuous(limits = c(-max_abs_logOR, max_abs_logOR), -->
<!--                      breaks = c(-4, -2, 0, 2, 4)) + # Symmetrical limits based on the maximum observed value -->
<!--     theme(axis.text.x = element_text(angle = 0, hjust = 1, vjust = 0.5, size = 10, color = "black"), -->
<!--         axis.text.y = element_text(color = "black"), -->
<!--         axis.title = element_text(color = "black"), -->
<!--         legend.text = element_text(color = "black"), -->
<!--         legend.title = element_text(color = "black"), -->
<!--         plot.title = element_text(color = "black"), -->
<!--         plot.subtitle = element_text(color = "black"), -->
<!--         plot.caption = element_text(color = "black"), -->
<!--     #     legend.position = "right",  # Set legend position to top -->
<!--     # legend.direction = "horizontal" -->
<!--     )  # Set legend direction to horizontal) -->
<!-- volcano_plot -->

<!-- # Save the plot -->
<!-- ggsave(here("Samples538_A_Geno_Microbiome/A3_AIC_BIC_comp/", -->
<!--             paste0( filext, "volcano_plot.png")), -->
<!--        plot = volcano_plot, -->
<!--        width = 9, height = 7, dpi = 300) -->
<!-- ``` -->


<!-- ## write the file with formatted values in a word table -->

<!-- ```{r} -->
<!-- Samples538_Plink_logit_covASPB_w_anno %>% head() -->

<!-- sel_cols = c("Location_GRCh38", "Gene", "SNP_rsID", "A1", "A2", "Frq_ECC", -->
<!--              "Frq_CF", "OR", "SE", "P", "P_Bonferroni", "Variant_Type") -->

<!-- Samples538_Plink_logit_covASPB_w_anno_2 = Samples538_Plink_logit_covASPB_w_anno %>%  -->
<!--   filter(P < 0.001) %>%  -->
<!--     mutate( -->
<!--     OR = ifelse(OR < 0.001, "<0.001", formatC(OR, format = "f", digits = 3)), -->
<!--     Frq_ECC = ifelse(Frq_ECC < 0.001, "<0.001", formatC(Frq_ECC, format = "f", digits = 3)), -->
<!--     Frq_CF = ifelse(Frq_CF < 0.001, "<0.001", formatC(Frq_CF, format = "f", digits = 3)), -->
<!--     P = formatC(P, format = "e", digits = 2), -->
<!--     P_Bonferroni = formatC(P_Bonferroni, format = "e", digits = 2)  -->
<!--     ) %>%  -->
<!--   ## remove _ from the Variant_Type -->
<!--   mutate(Variant_Type = gsub("_", " ", Variant_Type)) %>% -->
<!--   dplyr::select(sel_cols) -->

<!-- Samples538_Plink_logit_covASPB_w_anno_2 -->

<!-- ``` -->

<!-- ```{r} -->

<!-- # Order your data frame by GENE -->
<!-- Samples538_Plink_logit_covASPB_w_anno_3 <- Samples538_Plink_logit_covASPB_w_anno_2 %>% -->
<!--  group_by(Gene) %>% -->
<!--   mutate(min_P = min(as.numeric(P), na.rm = TRUE)) %>%  # Calculate the minimum P-value for each gene -->
<!--   arrange(min_P, Gene) %>%  # Arrange by minimum P-value first, then by gene -->
<!--   ungroup() %>%  # Ungroup the data if needed -->
<!--   select(-min_P)  # Remove the temporary min_P column if you don't need it -->
<!-- Samples538_Plink_logit_covASPB_w_anno_3 -->

<!-- # Define the width of each column with explicit column names -->
<!-- column_widths <- c( -->
<!--   Location_GRCh38 = 1.2, -->
<!--   Gene = 0.8, -->
<!--   SNP_rsID = 1, -->
<!--   A1 = 0.5, -->
<!--   A2 = 0.6, -->
<!--   Frq_ECC = 0.7, -->
<!--   Frq_CF = 0.7, -->
<!--   OR = 0.6, -->
<!--   SE = 0.7, -->
<!--   P = 0.9, -->
<!--   P_Bonferroni = 0.9, -->
<!--   Variant_Type = 1.5 -->
<!-- ) -->

<!-- # Create the flextable -->
<!-- ft <- flextable::regulartable(Samples538_Plink_logit_covASPB_w_anno_3) %>%  -->
<!--   flextable::width(j = names(column_widths), width = column_widths, unit = "in") -->

<!-- # Rest of your code remains the same -->
<!-- last_occurrence <- !duplicated(Samples538_Plink_logit_covASPB_w_anno_3$Gene, fromLast = TRUE) -->
<!-- ft <- flextable::border(ft, part = "body", border.bottom = officer::fp_border(color = "black"), i = which(last_occurrence)) -->

<!-- ft <- flextable::fontsize(ft, size = 8, part = "all") -->
<!-- ft <- flextable::merge_v(ft, j = "Gene") -->

<!-- # Add the FlexTable to a Word document in landscape orientation -->
<!-- doc <- officer::read_docx() %>% -->
<!--   flextable::body_add_flextable(value = ft, split = TRUE) %>% -->
<!--   officer::body_end_section_landscape() -->

<!-- # Save the Word document -->
<!-- print(doc, -->
<!--       target = here("Samples538_A_Geno_Microbiome/A3_AIC_BIC_comp", -->
<!--                     paste0(filext, "Variant_Table.docx")) -->
<!-- ) -->

<!-- ``` -->


