---
title: "mbQTL_sig_variant_all_species_dominant_model_NO_PCs"
output: html_document
date: "2024-08-20"

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r}
rm(list = ls())

```

# Load libraries
```{r}
library(tidyverse)
library(here)
library(snpStats)
library(vcfR)
library(mbQTL)
library(RColorBrewer)
```


## Load the variant file
```{r}
Samples538_Plink_logit_covASPB_w_anno <- 
  read.table(here("Samples538_A_Geno_Microbiome/A1_Plink/A1_02_Plink_dominant",
                  "A1_02_Samples538_55gene_logistic_dominant_cov_anno.txt"),
                                                          header = T)
Samples538_Plink_logit_covASPB_w_anno %>% head()
```


```{r}
Samples538_Plink_logit_covASPB_w_anno_2 = Samples538_Plink_logit_covASPB_w_anno %>% 
  filter(P_BH < 0.05) %>%
  filter(A1 != '*') %>%
  mutate(SNP_A1 = paste(SNP, A1, sep = "_"))
Samples538_Plink_logit_covASPB_w_anno_2 %>% head()
```







## load recoded vcf file to make a new vcf which as only the above variants
```{r}
Samples538_recoded_vcf <- 
  read.table(here("Samples538_A_Geno_Microbiome/A1_Plink/A1_00_Plink_files",
                  "Samples538_55gene_Plink.vcf.raw"),
                        header = T)
Samples538_recoded_vcf[, 1:10]  %>% head()

Samples538_recoded_vcf_2 = Samples538_recoded_vcf %>%
  column_to_rownames(var = "IID") %>%
  # select only the variants in the assoc file
  dplyr::select(Samples538_Plink_logit_covASPB_w_anno_2$SNP_A1) %>% 
  # replace missing values with 0
  mutate(across(everything(), ~if_else(is.na(.), 0, .))) %>% 
  # to make it a dominant vcf table
  mutate(across(everything(), ~case_when(
  . == 2 ~ 1,  # Replace 2 with 1
  TRUE ~ .     # Keep all other values the same
  )))

Samples538_recoded_vcf_2 %>% head()
```



## CLR normalized microbiome file with only significant species

```{r}

ALL_16S_ITS_features <- 
  read.table(here(
    "../RP_230718_Taste_Genes_Analysis_1/Samples538_Analysis/Microbiome538_files",
                  "ALL_16S_ITS_DA_normalized_features.txt"),
                                       header = T)
ALL_16S_ITS_features %>% head()

ALL_16S_ITS_features_2 = ALL_16S_ITS_features %>%
  column_to_rownames(var = "IID") %>%
  dplyr::select(-FID)
ALL_16S_ITS_features_2 %>% head()
```

##covariate

## Load the metadata file
```{r}
Samples538_covariates <- 
  read.table(here("Samples538_A_Geno_Microbiome/A1_Plink/A1_00_Plink_files",
                           "Samples538_Combined_covariates_PCA.txt"),
                      header = T)
Samples538_covariates %>% head()

# Samples538_covariates_2 = Samples538_covariates %>%
#   mutate(Batch = case_when(
#     Batch_2 == 0 & Batch_3 == 0 ~ 1,  # If Batch_2 = 0 and Batch_3 = 0, assign Batch = 1
#     Batch_2 == 1 & Batch_3 == 0 ~ 2,  # If Batch_2 = 1 and Batch_3 = 0, assign Batch = 2
#     Batch_2 == 0 & Batch_3 == 1 ~ 3   # If Batch_2 = 0 and Batch_3 = 1, assign Batch = 3
#   )) %>% 
#   mutate(Batch = as.factor(Batch)) %>%
#   mutate(ECC_status = case_when(
#     ECC_status == 0 ~ "CF",
#     ECC_status == 1 ~ "ECC"
#   )) %>% 
#   mutate(ECC_status = as.factor(ECC_status)) %>%
#   mutate(Sex =  case_when( # # AS 	required by plink Sex code '1' = male, '2' = female, '0' = unknown)
#     Sex == 1 ~ "Male",
#     Sex == 2 ~ "Female"
#   )) %>% 
#   mutate(Sex = as.factor(Sex))
# Samples538_covariates_2
```


```{r}
Samples538_covariates %>% group_by(Sex) %>% summarise(n = n())
Samples538_covariates %>% group_by(ECC_status) %>% summarise(n = n())
```

Required by plink Binary ('0' = control, '1' = case)
Required by plink Sex code ('1' = male, '2' = female, '0' = unknown)
FID (Family ID) and IID (Individual ID) for it to correctly identify individuals
mbQTL can not take non numerical covariates

```{r}
Samples538_covariates_2 = Samples538_covariates %>% 
  dplyr::select(IID, ECC_status, Sex, Age, Urban_status, SEFI_score, PC1, PC2, PC3, PC4, PC5) %>%
  # mutate(FID = IID) %>%
  mutate(Sex = case_when( # AS Original in data 1, Female | 2, Male 
    Sex == 0 ~ "Female",
    Sex == 1 ~ "Male")) %>% 
  mutate(Sex = case_when( # AS 	required by plink Sex code ('1' = male, '2' = female, '0' = unknown)
    Sex == "Female" ~ 2,
    Sex == "Male" ~ 1,
    TRUE ~ 0 )) %>% 
  dplyr::select(IID, ECC_status, Sex, Age, Urban_status, SEFI_score) %>% #, PC1, PC2, PC3, PC4, PC5) %>%
  column_to_rownames(var = "IID")

Samples538_covariates_2 %>% head()

Samples538_covariates_2_t = Samples538_covariates_2 %>%
  as.matrix() %>% 
  t() %>% 
  as.data.frame()
Samples538_covariates_2_t 
```



```{r}
Samples538_LinearAnalysisTaxaSNP <- linearTaxaSnp(ALL_16S_ITS_features_2,
  Samples538_recoded_vcf_2,
  Samples538_covariates_2_t
)
Samples538_LinearAnalysisTaxaSNP 

```



```{r}
Samples538_LinearAnalysisTaxaSNP_2 = Samples538_LinearAnalysisTaxaSNP %>%
  dplyr::select(Genus, snps, FDR) %>%
  mutate(P_BH_adj = as.numeric(FDR)) %>%
filter(P_BH_adj <= 0.05) %>% 
  mutate(nlog10P_BH_adj = -log10(P_BH_adj)) %>% 
  select(-FDR)
Samples538_LinearAnalysisTaxaSNP_2 
```
## Save as tsv
```{r}
# write.table(Samples538_LinearAnalysisTaxaSNP_2, 
#            file = here("Samples538_B_MicrbiomeQTL/B3_mbQTL",
#                        "B3_05_Samples538_mbQTL_all_variant_all_species.tsv"),
#            quote = F, sep = "\t", row.names = F)

```


```{r}
Samples538_LinearAnalysisTaxaSNP_spread = Samples538_LinearAnalysisTaxaSNP %>% # Samples538_LinearAnalysisTaxaSNP_2
    mutate(pvalue = as.numeric(pvalue)) %>%
  mutate(nlog10P = -log10(pvalue)) %>%
  # filter(nlog10P > 3) %>%
  dplyr::select(Genus, snps, nlog10P) %>%
  tidyr::spread(key = snps, value = nlog10P) %>% 
  # mutate_all(as.numeric) %>%
  replace(is.na(.), 0)
Samples538_LinearAnalysisTaxaSNP_spread %>% head()

Samples538_LinearAnalysisTaxaSNP_gather = Samples538_LinearAnalysisTaxaSNP_spread %>%
  gather(key = "snps", value = "nlog10P", -Genus) 
Samples538_LinearAnalysisTaxaSNP_gather %>% head()

Samples538_LinearAnalysisTaxaSNP_gather$Genus <- abbreviate(Samples538_LinearAnalysisTaxaSNP_gather$Genus, minlength = 35, use.classes = FALSE)
Samples538_LinearAnalysisTaxaSNP_gather %>% head()
```


```{r}
# Create the heatmap
Samples538_LinearAnalysisTaxaSNP_heatmap <- 
  ggplot(Samples538_LinearAnalysisTaxaSNP_gather,
         aes(y = snps, x = Genus, fill = nlog10P)) +
  geom_tile(color = "black") +
  scale_fill_gradientn(colors = c("white", "yellow", "green", "darkgreen", "darkred"),
                       limits = c(0, max(Samples538_LinearAnalysisTaxaSNP_gather$nlog10P, na.rm = TRUE))) +
  # scale_x_discrete(limits = rev(levels(ld_long$Var1))) +  # Reverse x-axis order
  scale_y_discrete(limits = rev(levels(Samples538_LinearAnalysisTaxaSNP_gather$snps)),
                   position = "right") +  # Move y-axis labels to the right
  theme_minimal() +
  theme(axis.text.x = element_text(size = 8, family = "Arial", color = "black",
                                   angle = 60, hjust = 1, vjust = 1),
        axis.text.y = element_text(size = 8, family = "Arial", color = "black"),
        axis.title.x = element_text(size = 9, face = "bold"),
        axis.title.y = element_text(size = 9, face = "bold"),
        legend.position = "left",  # Move the legend to the left
        legend.title = element_text(size = 8, face = "bold"),
        legend.text = element_text(size = 8)) +
  labs( y = "Genetic Variants", x = "Microbial Species",
        title = "mbQTL_sig_variant_all_species_dominant_model",
        subtitle =  "P < 0.05")

Samples538_LinearAnalysisTaxaSNP_heatmap
ggsave(here("Samples538_B_MicrbiomeQTL/B3_mbQTL_dominant",
            "B3_05.2_mbQTL_sig_variant_all_species_dominant_model.png"),
       Samples538_LinearAnalysisTaxaSNP_heatmap,
       width = 12,
       height = 4,
       dpi = 300)

```

```{r}
Samples538_LinearAnalysisTaxaSNP_2_spread = Samples538_LinearAnalysisTaxaSNP_2 %>% # Samples538_LinearAnalysisTaxaSNP_2_2
  #   mutate(FDR = as.numeric(FDR)) %>%
  # mutate(nlog10P_BH = -log10(FDR)) %>%
  filter(nlog10P_BH_adj > 1) %>%
  dplyr::select(Genus, snps, nlog10P_BH_adj) %>%
  tidyr::spread(key = snps, value = nlog10P_BH_adj) %>% 
  # mutate_all(as.numeric) %>%
  replace(is.na(.), 0)
Samples538_LinearAnalysisTaxaSNP_2_spread %>% head()

Samples538_LinearAnalysisTaxaSNP_2_gather = Samples538_LinearAnalysisTaxaSNP_2_spread %>%
  gather(key = "snps", value = "nlog10P_BH_adj", -Genus) 
Samples538_LinearAnalysisTaxaSNP_2_gather %>% head()

Samples538_LinearAnalysisTaxaSNP_2_gather$Genus <- abbreviate(Samples538_LinearAnalysisTaxaSNP_2_gather$Genus, minlength = 35, use.classes = FALSE)
Samples538_LinearAnalysisTaxaSNP_2_gather %>% head()
```


```{r}
# Create the heatmap
Samples538_LinearAnalysisTaxaSNP_2_heatmap <- 
  ggplot(Samples538_LinearAnalysisTaxaSNP_2_gather,
         aes(y = snps, x = Genus, fill = nlog10P_BH_adj)) +
  geom_tile(color = "black") +
  scale_fill_gradientn(colors = c("white", "yellow", "green", "darkgreen", "darkred"),
                       limits = c(0, max(Samples538_LinearAnalysisTaxaSNP_2_gather$nlog10P_BH_adj, na.rm = TRUE))) +
  # scale_x_discrete(limits = rev(levels(ld_long$Var1))) +  # Reverse x-axis order
  scale_y_discrete(limits = rev(levels(Samples538_LinearAnalysisTaxaSNP_2_gather$snps)),
                   position = "right") +  # Move y-axis labels to the right
  theme_minimal() +
  theme(axis.text.x = element_text(size = 8, family = "Arial", color = "black",
                                   angle = 60, hjust = 1, vjust = 1),
        axis.text.y = element_text(size = 8, family = "Arial", color = "black"),
        axis.title.x = element_text(size = 9, face = "bold"),
        axis.title.y = element_text(size = 9, face = "bold"),
        legend.position = "left",  # Move the legend to the left
        legend.title = element_text(size = 8, face = "bold"),
        legend.text = element_text(size = 8)) +
  labs( y = "Genetic Variants", x = "Microbial Species",
        title = "mbQTL_sig_variant_all_species_dominant_model_NO_PCs",
        subtitle =  "nlog10P_BH < 0.05")

Samples538_LinearAnalysisTaxaSNP_2_heatmap
ggsave(here("Samples538_B_MicrbiomeQTL/B3_mbQTL_dominant",
            "B3_05.2_mbQTL_sig_variant_all_species_dominant_model_P_BH.png"),
       Samples538_LinearAnalysisTaxaSNP_2_heatmap,
       width = 12,
       height = 4,
       dpi = 300)

```
