---
title: "B4_02_Species_sig_variant_all_OTU lm"
output: html_document
date: "2024-11-09"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r}
rm(list = ls())
```

```{r}
file_name_ext="B4_02_Species_sig_variant_all_OTU"
```

# Load libraries
```{r}
library(tidyverse)
library(here)
library(LDM)
library(ComplexHeatmap)
library(circlize)
library(grid)
library(cowplot)
library(patchwork)
```


## Need three tables
1. A microbiome table

16S microbiome
#### For microbiome data use Maaslin features for consistency in naming
```{r}
# Import the Maaslin features
Plaque538_16S_Species_CLR = readr::read_tsv(here("Samples538_A_Geno_Microbiome/A2_Maaslin/Maaslin2_16S_Species_output/features",
                                            "filtered_data_norm.tsv"))
Plaque538_16S_Species_CLR
```
ITS microbiome
```{r}
# Import the Maaslin features
Plaque538_ITS_Species_CLR = readr::read_tsv(here("Samples538_A_Geno_Microbiome/A2_Maaslin/Maaslin2_ITS_Species_output/features",
                                            "filtered_data_norm.tsv"))
Plaque538_ITS_Species_CLR

```


```{r}

Plaque538_16S_ITS_Species_CLR_OTU = full_join(Plaque538_16S_Species_CLR, Plaque538_ITS_Species_CLR, by = "feature") %>%
  rename(IID = feature)
Plaque538_16S_ITS_Species_CLR_OTU %>% head()

# Plaque538_16S_ITS_Species_CLR_OTU_2 = Plaque538_16S_ITS_Species_CLR_OTU %>%
#   # column_to_rownames(var = "IID") %>%
#   # dplyr::select(-FID) %>% 
#   identity()
# Plaque538_16S_ITS_Species_CLR_OTU_2 

Plaque538_16S_ITS_Species_CLR_OTU_names = Plaque538_16S_ITS_Species_CLR_OTU %>% 
  dplyr::select(-IID) %>%
  colnames()
Plaque538_16S_ITS_Species_CLR_OTU_names
```



## microbiome file with only Maaslin2 significant species

```{r}
Maaslin2_16S_ITS_features =
  readRDS(here("Samples538_A_Geno_Microbiome/A2_Maaslin",
               "A2_04_Species_coef_qval_16S_ITS.rds"))
Maaslin2_16S_ITS_features

# Maaslin2_16S_ITS_features_2 = Maaslin2_16S_ITS_features %>%
#   # column_to_rownames(var = "IID") %>%
#   dplyr::select(-FID) %>%
#   # as.matrix() %>%
#   # filter(IID %in% rownames(Samples538_covariates_vcf_raw)) %>%
#   column_to_rownames(var = "IID") %>%
#   identity()
# Maaslin2_16S_ITS_features_2 %>% head()

Maaslin2_16S_ITS_features_names = Maaslin2_16S_ITS_features %>%
  filter(q_values < 0.01) %>%
  pull(OTUs)
Maaslin2_16S_ITS_features_names

```




```{r}
# Plaque538_16S_ITS_Species_CLR_OTU_sig = Plaque538_16S_ITS_Species_CLR_OTU_2 %>%
#   # dplyr::select(one_of(Maaslin2_16S_ITS_features_names)) %>% 
#   identity()
# Plaque538_16S_ITS_Species_CLR_OTU_sig
# 
# ## feature names
# Plaque538_16S_ITS_Species_CLR_OTU_sig_names = Plaque538_16S_ITS_Species_CLR_OTU_sig %>% 
#   dplyr::select(-IID) %>%
#   names()
# Plaque538_16S_ITS_Species_CLR_OTU_sig_names 

```




## load recoded vcf file to make a new vcf which as only the above variants
```{r}
Samples538_recoded_vcf <- 
  read.table(here("Samples538_A_Geno_Microbiome/A1_Plink/A1_00_Plink_files",
                  "Samples538_55gene_Plink.vcf.2.raw"),
                        header = T)
Samples538_recoded_vcf[, 1:10]  %>% head()


# Create a vector of column names
columns_to_remove <- c("FID", "PAT", "MAT", "SEX", "PHENOTYPE")


Samples538_recoded_vcf_2 = Samples538_recoded_vcf %>%
  column_to_rownames(var = "IID") %>%
  dplyr::select(-c(columns_to_remove)) %>%
  # replace missing values with 0
  mutate(across(everything(), ~if_else(is.na(.), 0, .)))  %>% 
  # to make it a dominant vcf table
  mutate(across(everything(), ~case_when(
  . == 2 ~ 1,  # Replace 2 with 1
  TRUE ~ .     # Keep all other values the same
  ))) %>%
  # change the variants to factor
  mutate(across(everything(), as.factor)) %>%
  ## IID column is required for merging
  # rownames_to_column(var = "IID") %>% 
  identity()

Samples538_recoded_vcf_2 %>% head()
```
## Load the variant file
```{r}
Samples538_Plink_logit_covASPB_w_anno <- 
  read.table(here("Samples538_A_Geno_Microbiome/A1_Plink/A1_02_Plink_dominant",
                  "A1_02_Samples538_55gene_logistic_dominant_cov_anno.txt"),
                                                          header = T)
Samples538_Plink_logit_covASPB_w_anno %>% head()

Samples538_Plink_logit_covASPB_w_anno_2 = Samples538_Plink_logit_covASPB_w_anno %>% 
  arrange(P_BH) %>%                 # Sort by P_BH in ascending order
  slice_head(prop = 0.05) %>%          # Select the top 1% of rows
  # filter(P_BH < 0.05) %>%
  filter(A1 != '*') %>%
  # mutate(SNP_A1 = paste(SNP, A1, sep = "_")) %>% 
  identity()
Samples538_Plink_logit_covASPB_w_anno_2 

Samples538_Plink_logit_covASPB_w_anno_2_sig = Samples538_Plink_logit_covASPB_w_anno_2 %>%
  filter(P_BH < 0.05) %>%
  pull(SNP)
Samples538_Plink_logit_covASPB_w_anno_2_sig
```

```{r}
Samples538_recoded_vcf_sig = Samples538_recoded_vcf_2 %>%
  dplyr::select(Samples538_Plink_logit_covASPB_w_anno_2$SNP) %>% 
  rownames_to_column(var = "IID")
Samples538_recoded_vcf_sig

## variant names
Samples538_recoded_vcf_sig_names = Samples538_recoded_vcf_sig %>% 
  dplyr::select(-IID) %>%
  names()
Samples538_recoded_vcf_sig_names 
```



##covariate

## Load the metadata file
```{r}
Samples538_covariates <- 
  read.table(here("Samples538_A_Geno_Microbiome/A1_Plink/A1_00_Plink_files",
                           "Samples538_Combined_covariates_PCA.txt"),
                      header = T)
Samples538_covariates %>% head()


```


```{r}
Samples538_covariates %>% group_by(Sex) %>% summarise(n = n())
Samples538_covariates %>% group_by(ECC_status) %>% summarise(n = n())
```

Required by plink Binary ('0' = control, '1' = case)
Required by plink Sex code ('1' = male, '2' = female, '0' = unknown)
FID (Family ID) and IID (Individual ID) for it to correctly identify individuals
mbQTL can not take non numerical covariates

```{r}

covariates <- c("Sex", "Age", "RuralUrban", "SEFI_score", "PC1", "PC2", "PC3", "PC4", "PC5")

Samples538_covariates_2 = Samples538_covariates %>% 
  dplyr::select(IID, ECC_status, Sex, Age, RuralUrban, SEFI_score, PC1, PC2, PC3, PC4, PC5) %>%
  # mutate(FID = IID) %>%
  mutate(Sex = case_when( # AS Original in data 1, Female | 2, Male 
    Sex == 0 ~ "Female",
    Sex == 1 ~ "Male")) %>% 
  mutate(Sex = case_when( # AS 	required by plink Sex code ('1' = male, '2' = female, '0' = unknown)
    Sex == "Female" ~ 2,
    Sex == "Male" ~ 1,
    TRUE ~ 0 )) %>% 
  dplyr::select(IID, ECC_status, covariates) %>%
  mutate(across(c(ECC_status, Sex, RuralUrban), as.factor)) %>% 
  # column_to_rownames(var = "IID") %>% 
  identity()

Samples538_covariates_2 


# Samples538_covariates_2_t = Samples538_covariates_2 %>%
#   as.matrix() %>% 
#   t() %>% 
#   as.data.frame()
# Samples538_covariates_2_t 
```










# combine all the datasets


```{r}
# Combine datasets by joining on "IID" and applying transformations in a streamlined manner
Samples538_covariates_vcf_clr_microbiome <- Samples538_covariates_2 %>%
  inner_join(Samples538_recoded_vcf_sig, by = "IID") %>%
  inner_join(Plaque538_16S_ITS_Species_CLR_OTU, by = "IID") %>%
  # Remove rows with NA values
  na.omit() %>%
  identity()

Samples538_covariates_vcf_clr_microbiome
```




```{r}
library(mediation)
linear_model <- glm(B_s_Streptococcus_mutans ~ chr15_40303364_PLCB2_rs2305645 +Age+Sex+PC1+PC2+PC3+PC4+PC5+RuralUrban+SEFI_score,
                     data =Samples538_covariates_vcf_clr_microbiome,
                     family = gaussian)

# outcome_model <- glm(ECC_status ~ chr15_40303364_PLCB2_rs2305645_T + B_Streptococcus_mutans +Age+Sex+PC1+PC2+PC3+PC4+PC5+RuralUrban+SEFI_score,
#                     data = Samples538_covariates_vcf_clr_microbiome,
#                     family = binomial(link = "logit"))
# 
# results <- mediate(mediator_model,
#                    outcome_model,
#                    covariates = c("Age", "Sex", "PC1", "PC2", "PC3", "PC4", "PC5", "RuralUrban", "SEFI_score"),
#                    treat= 'chr15_40303364_PLCB2_rs2305645_T',
#                    mediator= 'B_Streptococcus_mutans' ,
#                    boot=TRUE,
#                    sims=5)
# results %>% summary()
linear_model %>% broom::tidy()
```

```{r}
```

```{r}

summary(linear_model)$coef[2, "Estimate"]
```

```{r}
set.seed(4321)
# Create an empty dataframe to store the results
linear_model_results_df <- data.frame(
  variant = character(),
  OTU = character(),
  linear_model_estimate = numeric(),
  linear_model_p_value = numeric(),
  stringsAsFactors = FALSE
)


# Iterate over each combination of variant and OTU
for (OTU in Plaque538_16S_ITS_Species_CLR_OTU_names) {
  for (variant in Samples538_recoded_vcf_sig_names) {

    # Fit the mediator model
    linear_model <- glm(as.formula(paste0(OTU, " ~ ", variant,
                                            " + Age + Sex + RuralUrban + SEFI_score + PC1 + PC2 + PC3 + PC4 + PC5 + ECC_status")),
                          data = Samples538_covariates_vcf_clr_microbiome,
                          family = gaussian)

    # Extract ACME (average) estimate and p-value
    linear_model_estimate <- summary(linear_model)$coef[2, "Estimate"]
    linear_model_p_value <- summary(linear_model)$coef[2, "Pr(>|t|)"]
    # Store the results in the dataframe
    linear_model_results_df <- linear_model_results_df %>%
      add_row(variant = variant,
              OTU = OTU,
              linear_model_estimate = linear_model_estimate,
              linear_model_p_value = linear_model_p_value)
  }
}
linear_model_results_df
```


```{r}
# Display the results dataframe
linear_model_results_df_2 = linear_model_results_df %>% 
  # apply BH correction to p-values
  mutate(linear_model_p_BH = p.adjust(linear_model_p_value, method = "BH")) %>%
  filter(linear_model_p_value < 0.001) %>% 
  arrange(linear_model_p_value) %>%
  mutate(variant = factor(variant, levels = Samples538_recoded_vcf_sig_names)) %>% 
  identity()
linear_model_results_df_2
```

```{r}
# Create new columns for p-value significance and FDR significance
linear_model_results_df_3 <- linear_model_results_df_2 %>%
  complete(variant, OTU,
           fill = list(linear_model_estimate = 0,
                       linear_model_p_value = NA,
                       linear_model_p_BH = NA)) %>%
  mutate(p_significance = case_when(
    linear_model_p_value < 0.00001 ~ "*****",
    linear_model_p_value < 0.0001 ~ "****",
    linear_model_p_value < 0.001 ~ "***",
    linear_model_p_value < 0.01  ~ "**",
    linear_model_p_value < 0.05  ~ "*",
    TRUE            ~ ""
  ),
  fdr_significance = case_when(
    linear_model_p_BH < 0.05 ~ "†",
    TRUE             ~ ""
  )) %>% 
  mutate(significance = paste0(p_significance, fdr_significance))
linear_model_results_df_3
```
```{r}
# Make a modified dataframe for custom axis styling
linear_model_results_df_3 <- linear_model_results_df_3 %>%
  mutate(
    x_face = ifelse(variant %in% Samples538_Plink_logit_covASPB_w_anno_2_sig,
                    paste0("**", linear_model_results_df_3$variant, "**"), # Bold formatting
                        linear_model_results_df_3$variant),
    y_face = ifelse(OTU %in% Maaslin2_16S_ITS_features_names, "bold", "plain")
  )
linear_model_results_df_3
# Create custom labels: bold for categories in `bold_categories`, plain otherwise
custom_labels <- ifelse(linear_model_results_df_3$variant %in% Samples538_Plink_logit_covASPB_w_anno_2_sig,
                        paste0("**", linear_model_results_df_3$variant, "**"), # Bold formatting
                        linear_model_results_df_3$variant)                    # Plain text
```


```{r}
variant_otu_heatmap =
# Adjust labels separately
ggplot(linear_model_results_df_3, aes(x = variant, y = OTU, fill = linear_model_estimate)) +
  geom_tile(color = "black", size = 0.1) +
  geom_text(aes(label = significance), vjust = 0.5, color = "black", size = 2) +
  scale_fill_gradient2(low = "cadetblue4", mid = "white", high = "chocolate4", 
                       midpoint = 0, name = "Estimate") +
  scale_y_discrete(labels = function(labels) str_trunc(labels, width = 35, side = "right")) +
   scale_x_discrete(labels = function(labels) str_replace(labels, "chr\\d+[_]+\\d+_", "")) +
  theme_minimal(base_size = 12) +
  labs(x = "Genetic variants", y = "Microbial species") +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1.1
                               , vjust = 1.15
                               , size = 6, color = "black" ),
    axis.text.y = element_text(size = 6, color = "black"),
    axis.title = element_text(size = 7, color = "black", face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid = element_blank(),
    legend.position = "top",
    legend.key.size = unit(0.25, "cm"),
    legend.text = element_text(size = 4),
    legend.title = element_text(size = 6)
  ) #+
variant_otu_heatmap
  # scale_x_discrete(labels = linear_model_results_df_3$x_face) # Apply the dynamically created labels
ggsave(here("Samples538_B_MicrbiomeQTL/B4_lm",
         paste0(file_name_ext, "_Heatmap.jpg")),
       variant_otu_heatmap,
       width = 6,
       height = 5,
       units = "in")
# Save the combined plot

```

# Combine it with the Maaslin2 coefficients


# filter the OUTs identified with variant association
```{r}
Maaslin2_16S_ITS_features_coef = Maaslin2_16S_ITS_features %>%
  filter(OTUs %in% linear_model_results_df_3$OTU)
Maaslin2_16S_ITS_features_coef
```

# plot forest type plot for the Maaslin2 coefficients
```{r}
Maaaslin2_16S_ITS_features_coef_plot =
# Create the forest plot with p-value annotations
ggplot(Maaslin2_16S_ITS_features_coef, aes(x = Coefficients, y = OTUs)) +
  geom_point(color = "black", size = 1) +
  scale_y_discrete(labels = function(labels) str_trunc(labels, width = 35, side = "right")) +
  geom_errorbarh(aes(xmin = lower_ci, xmax = upper_ci), height = 0.2, color = "black", linewidth = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", size = 0.2) +
  # geom_text(aes(label = sprintf("q = %.3f", q_values)), 
  #           hjust = 0, vjust = 1.3, size = 3, color = "black") +
  theme_minimal(base_size = 12) +
  labs(
    title = NULL,
    x = "MaAsLin2 Coefficient",
    y = "Microbial species"
  ) +
  theme(
    
    axis.text.x = element_text(size = 6, color = "black"),
    axis.text.y = element_text(size = 6, color = "black"),
    axis.title = element_text(size = 7, color = "black", face = "bold"),
  )
Maaaslin2_16S_ITS_features_coef_plot
```
## Combine the two plots using patchwork

```{r}
# Combine the two plots with patchwork
combined_plot <-  
  variant_otu_heatmap + 
  Maaaslin2_16S_ITS_features_coef_plot + plot_layout(axes = "collect")

# Display the combined plot
print(combined_plot)

# Save the combined plot
ggsave(here("Samples538_B_MicrbiomeQTL/B4_lm",
         paste0(file_name_ext, "_Heatmap_Forest.jpg")),
       combined_plot,
       width = 14,
       height = 10,
       units = "cm")
```




